Thank you, next: demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2024-04-26 Manuscript/ant_cactus_AmNat_v2.tex
Manuscript

	\section*{Abstract}
% 200 words max
The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.
\textit{Cylindropuntia imbracata} (tree cholla) produce extrafloral nectar and various ant partners provide defense from herbivores and seed predators. 
We used plant demographic censuses to parameterize a series of Bayesian hierarchical generalized linear vital rate models to determine the impacts of different partners on the focal mutualists. 
We constructed an Integral Projection Model in which we simulate different combinations of ant partners that don’t occur in the wild.
The hierarchical models revealed that different ant partners had different impacts on the cholla vital rates. 
Specifically, \textit{C. opuntiae} tended plants had advantages in both growth and survival when small, and large \textit{L. apiculatum} tended plants had floral viability advantages. 
The IPM results revealed that scenarios which included \textit{L. apiculatum} resulted in the highest possible fitness for the tree cholla. 
This suggests that diversity benefits in this system are explained by sampling effect.
This study highlights that partner diversity can increase the overall benefits a focal mutualist receives and the importance of a mechanistic understanding to explain the benefits of this diversity across systems.

@@

@@

@@

\section*{Results}
\subsection*{What are the demographic effects of association with alternative parnters and how do these effects fluctuate across years?}
Over the 20-year data set, we found that ant partners influenced demographic performance of cactus hosts, and different ant partners had contrasting demoghraphic effects across host vital rates. 
\tom{Plants tended by \textit{C. opuntiae} had a growth advantage, particularly for small plants, while plants in states of \textit{L. apiculatum}, Other ants, and vacancy had indistinguishable growth trajectories (Figure \ref{fig:Grow}).}{Your draft had posterior probabilities associated with growth differences. I think it would be possble to keep the vital rate results very simple -- like what I have here -- and keep the posterior probability estimates for the IPM results. I am open to including posterior estimates for the vital rates but it is not clear to me how those values account for size, since we are comparing functions not single values. Let's discuss.} 
For all ant states, growth was left-skewed, with transitions to sizes below the mean were more common than sizes above the mean. 
Similarly, ant visitation enhances cactus survival, and ant partner identity has a significant impact on survival for smaller plants (Figure \ref{fig:Surv}).
Mean survival rates ranged from 7.7\% to 99.9\%, with the smallest plants the most vulnerable to mortality. 
\textit{C. opuntiae}-occupied plants had a survival advantage over other ant states, particularly at smaller sizes, consistent with the positive effects on growth. 
At larger sizes, plants in any state of ant occupancy had a survival advantage over vacant plants.

@@

Despite this apparent synergy, when partner identity was considered, we found the benefits of partner diversity could be explained by Sampling Effect.
We believe the benefits of partner diversity are heavily driven by the presence of a single best partner rather than overall synergy.
All simulated combinations of ant partners which included \textit{L. apiculatum} were nearly equal and the highest possible fitness estimated for the cholla. 
This indicates that \textit{L. apiculatum} are the single best partner for the cholla under existing conditions. 
Based on the definitions of Sampling Effect and Complementarity we use in this study \citep{Batstone2018}, it is clear that Sampling Effect can explain the benefits of partner diversity in the cholla system.

It is possible that this, like the frequency of partner turnover to \textit{L. apiculatum}, is driven by the extreme frequency of \textit{L. apiculatum} ants in comparison to others. 
With this in mind, we simulated the population fitness with equal probability for transitioning to any ant state.
We found ...... from the simulations with different transition probabilities.

@@

What now remains is to determine the underlying processes that lead to the emergence of these characterized patterns. 
Throughout the literature, it is reported that non-symbiotic ants and non-mutualist ants are highly attracted to monosaccharide heavy EFN \citep{Heil2004,Kautz2009} and that plants often produce monosaccharide EFN to avoid exploitation \citep{Heil2010}.
It has been previously reported that when reproducing, the nectar of stem segments has a greater disaccharide vs monosaccharide ratio, while flowerbuds have a lower ratio \citep{Miller2008}. 
This could potentially have a significant impact on the partners who tend a plant in a given season, though determining the impacts of nectar on directional turnover requires further study.

@@

% Limitations
As with any study, there are limitations to consider when interpreting the results reported here.
These results are based on observational data regarding ant effects on plant demography rather than experimental data, meaning we are able to determine correlations but not causation.
We are able to report evidence that plants experience higher growth when associating with a particular partner, but we are not able to determine if the presence of the ant is the cause of the higher growth observed.
It is possible, for example, that plants which attract \textit{L. apiculatum} through nectar composition have a different pattern of resource allocation to growth and survival and this is what we are observing rather than a direct affect of \textit{L. apiculatum} on the plant.
This is a typical limitation of observational studies, but does not reduce the power or importance of our results, just the interpretation.
However, we believe that the ants likely have direct impacts on the plant demography based on other studies which find the presence of ants can lead to direct increases in survival \citep{Beattie1985}, reproductive efforts \citep{Ford2015}, growth \citep{Frederickson2005}, etc.
In the future, it would be interesting to verify the causation of our results through experimental manipulations.
One further expansion could include nectar analysis of these cacti in conjuntion with ant interactions to look not only at the ant parnter impacts on the plant demography, but also at the ability of plants to attract specific partners through nectar composition shifts.

@@

% Conclusion

Discussion brainstorming:

1. Mini abstract paragraph
- synthesis of question plus answers, and overall significance
Topic sentence:

2. COnnecting to broader patterns in the literature
- address how complimentarity or synergies are commonly reported -- why did we not see this, especially since the vital rates suggest some complimentarity
- may involve the simulation experiments
- some contrast with other systems where benefits of diversity have been found
Topic sentence:

3. If there is one best partner then diversity should impose opportunity costs -- why do we not see this?
- should address how some systems can host multiple partners at a time while others (ours) can exclusive associations; for the latter diversity should only be costly if there is one best partner?

4. Turnover
- what's going on with turnover? what processes may drive the patters we see
- what is seen elsewhere in the lit (invertase papers may be an interesting reference? showing plant ``choice'')
- patterns could be consistent with ant and or plant choice (currently a ``black box'')
- this might address both frequency and direction (direction may lead to some discussion of competition)

5.  Portfolio effect
- what is it, what does it predict, what did we find
- what have other studies found? 
Topic sentence:

6. Value of long-term data
- includes turnover, portfolio effect
- also address how the full demographic data provides the correct insight into ant effects (eg We thought Crem was a parasite based on less complete data)

7. Herbivory
- this should underlie why these interactions are beneficial, so we need to address this
- this connects to portfolio effect, because fluctuations in herbivores (in conjunction with ant-specific defenses against herbivores) would be why porfolio effect could occur

8. Limitations
- confounds fundamental to observational data regarding ant effects on plant demography (correlation/causation problem) -- maybe plants that attract Liom through nectar have a different pattern of allocation o growth and survival
- cite experiments to support causal interpretation of observational patterns (ants really do stuff)
- extrapolating effects of ants beyond the interactions that we can actually see (e.g., new recruits with Liom); and the counterfactual of a Liom-only population is another form of extrapolation

9. Future directions -- maybe embed these into paragraphs above where we see opportunity to address limitations or extend the results -- probably lots we can do with herbivory

Conclusion.

Appendices:
1. Herbivory
2. Additional methods and results
3. Posterior predictive checks and model convergence
4. Maybe ant transition simulations

@@

\section*{Appendix A: Additional Methods and Parameters}

This is not referenced in the paper, to my knowledge, and I think you need to think more deeply about what content should go into appendices and why. 
% In most cases, authors should typeset supplementary material in a separate,
% author-supplied PDF. For author-supplied PDFs, please consult the
% AmNat_supp_template.tex document, available from
% https://www.journals.uchicago.edu/journals/an/instruct 
%
% By contrast, the Appendix instructions below apply to cases in which
% a brief appendix is to appear in print after the main body of the article.
% That notably includes descriptions of methods, tables defining parameters,
% and other material necessary for reproducing the MS's results.
%
% Please reset counters for the appendix (thus normally figure A1, 
% figure A2, table A1, etc.).
%
% Most AmNat articles have no more than one print appendix. If your article
% has more than one, counters for each appendix should match the letter of
% that appendix. For example, tables in Appendix B should be numbered table B1, % table C2, etc. This applies to tables, equations, and figures.
%
% It's better not to use the \appendix command, because we have some
% formatting peculiarities that \appendix conflicts with.

@@

\section*{Supplementary Materials}
\subsection*{Herbivory Data}
\subsection*{Model Checks}

>>>>

	\section*{Abstract}
% 200 words max
The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.
\textit{Cylindropuntia imbracata} (tree cholla) produce extrafloral nectar and various ant partners provide defense from herbivores and seed predators. 
We used plant demographic censuses to parameterize a series of Bayesian hierarchical generalized linear vital rate models to determine the impacts of different partners on the focal mutualists. 
We constructed an Integral Projection Model in which we simulate different combinations of ant partners that don’t occur in the wild.
The hierarchical models revealed that different ant partners had different impacts on the cholla vital rates. 
Specifically, \textit{C. opuntiae} tended plants had advantages in both growth and survival when small, and large \textit{L. apiculatum} tended plants had floral viability advantages. 
The IPM results revealed that scenario which included \textit{L. apiculatum} only resulted in the highest possible fitness for the tree cholla. 
This suggests that there are no benefits of diversity in this system.
This study highlights that partner diversity does not always increase the benefits a focal mutualist receives.

@@

We used this model to estimate probabilities which are counterfactual (such as the probability of being \textit{L. apiculatum} tended or vacant with no other options). 
To do this we assumed that the level of vacancy was constant across all states and that whatever ants were present in our simulation would tend the gaps which would otherwise be tended by another ant species. 
This assumes that competitive exclusion is the mechanism which determines the ants we see tending cacti, other potential approaches are outlined in Apendix XX.

@@

\subsubsection*{SIMULATION IDK WHAT TO CALL THIS RN }
Using an alternative transition model to parameterize the IPM, we conducted simulation experiments to determine if the sheer number of \textit{L. apiculatum} ants might be driving the outcomes of our partner diversity experiments. 
We followed the same protocol as the partner diversity experiments, except the transition model equalized the frequency of each ant partner.
We still preserved the observed pattern of size-dependent vacancy/occupancy, but occupancy was manipulated to be equally likely for all partner identities.

@@

\section*{Results}
\subsection*{What are the demographic effects of association with alternative parnters and how do these effects fluctuate across years?}
Over the 20-year data set, we found that ant partners influenced demographic performance of cactus hosts, and different ant partners had contrasting demoghraphic effects across host vital rates. 
Plants tended by \textit{C. opuntiae} had a growth advantage, particularly for small plants, while plants in states of \textit{L. apiculatum}, Other ants, and vacancy had indistinguishable growth trajectories (Figure \ref{fig:Grow}).
For all ant states, growth was left-skewed, with transitions to sizes below the mean were more common than sizes above the mean. 
Similarly, ant visitation enhances cactus survival, and ant partner identity has a significant impact on survival for smaller plants (Figure \ref{fig:Surv}).
Mean survival rates ranged from 7.7\% to 99.9\%, with the smallest plants the most vulnerable to mortality. 
\textit{C. opuntiae}-occupied plants had a survival advantage over other ant states, particularly at smaller sizes, consistent with the positive effects on growth. 
At larger sizes, plants in any state of ant occupancy had a survival advantage over vacant plants.

@@

Despite the positive effects of ant visitation in any combination of partners, when identity of partners is considered, we found that there is an opportunity cost to partner diversity.
There is no fitness scenario better than \textit{L. apiculatum} ants alone.
All other scenarios lead to diminished fitness of tree cholla, indicating that there is not a benefit of diversity in this system.
We originally believed the benefits of partner diversity are heavily driven by the presence of a single best partner rather than overall synergy.

After further simulations, we found that this lack of diversity benefits are not driven by frequency of a single ant. 
Using the simulations where all ants had equal frequencies across sizes, we found that scenarios with \textit{C. opuntiae} and Other ants present experienced fitness boosts under this simulation, though not eough to offer better fitness than \textit{L. apiculatum}.
Equal probability for transitioning into any ant state meant that the numbers of these ants were boosted significantly.

@@

What now remains is to determine the underlying processes that lead to the emergence of these characterized patterns. 
Throughout the literature, it is reported that non-symbiotic ants and non-mutualist ants are highly attracted to monosaccharide heavy EFN \citep{Heil2004,Kautz2009} and that plants often produce monosaccharide EFN to avoid exploitation \citep{Heil2010}.
It has been previously reported that when reproducing, the nectar of stem segments has a greater disaccharide vs monosaccharide ratio, while flowerbuds have a lower ratio \citep{Miller2008}. 
This could potentially have a significant impact on the partners who tend a plant in a given season, though determining the impacts of nectar on directional turnover requires further study. 
We conducted alternative simulations where the driving mechanism for ant occupation was not competitive exclusion, these results are included in Apendix XX.

@@

% Limitations
As with any study, there are limitations to consider when interpreting the results reported here.
These results are based on observational data regarding ant effects on plant demography rather than experimental data, meaning we are able to determine correlations but not causation.
We are able to report evidence that plants experience higher growth when associating with a particular partner, but we are not able to determine if the presence of the ant is the cause of the higher growth observed.
It is possible, for example, that plants which attract \textit{L. apiculatum} through osition have a different pattern of resource allocation to growth and survival and this is what we are observing rather than a direct affect of \textit{L. apiculatum} on the plant.
This is a typical limitation of observational studies, but does not reduce the power or importance of our results, just the interpretation.
However, we believe that the ants likely have direct impacts on the plant demography based on other studies which find the presence of ants can lead to direct increases in survival \citep{Beattie1985}, reproductive efforts \citep{Ford2015}, growth \citep{Frederickson2005}, etc.
In the future, it would be interesting to verify the causation of our results through experimental manipulations.
One further expansion could include nectar analysis of these cacti in conjuntion with ant interactions to look not only at the ant parnter impacts on the plant demography, but also at the ability of plants to attract specific partners through nectar composition shifts.

@@

@@

@@

\section*{Appendix A: Additional Methods and Parameters}
In addition to the models described in the body of the paper, we fit several other simpler models using data from previous studies.
These models are described below.

\section*{Appendix B: Observed Herbivory Data}
One type of data we took every year is herbivory data, which involved looking at the plant for evidence of recent herbivory (fresh bites) and noting the type and quantity of herbivores observed.

\section*{Appendix C: Alternative Ant Transition Simulations}
In addition to the competitive exclusion model and the equal probability (null) model defined and analyzed in the body of the paper, we simulated results from several other potential models. 
We chose to include competitive exclusion as our primary results in the paper because we believe it to be the most biologically realistic.
However, in the building and testing of alternative models we found that the method of ants occupying plants significantly impacts the fitness of the population.

\paragraph{Frequency Model}
The first alternative hypothesis we tested was what we called the frequncy model.
In this model rather than the proportion of vacant cacti being maintained, the proportion of cacti occupied by each species is maintained and when one is removed it is replaced with vacancy.
This version of the model assumes that the frequency of each ant we see is reflective of the real frequency of populations rather than some other mechanism.
With this model we found that both \textit{C. opuntiae} and Other ants are parasites when they are the only partner present (Figure \ref{fig:FreqLambdaMeans}a).
This indicates that at low frequencies, these ants are not able to provide strong benefits.
Diversity is beneficial through a sampling effect in this simulation (Figure \ref{fig:FreqLambdaMeans}b).

\begin{figure}
\includegraphics[width=0.61\linewidth]{Figures/lambda_means_freq.png}
\includegraphics[width=0.39\linewidth]{Figures/Lambda_Num_Partners_freq.png}
\caption{Panel a) shows the mean values of the estimated $\lambda_{S}$ (filled in circles) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{S}$ for }
\label{fig:FreqLambdaMeans}
\end{figure}

\paragraph{In-Between Model}
The first alternative hypothesis we tested was what we called the in-between model.
In this model rather than only competitive exclusion or frequency based occupancy, the occupancy is in-between these two models. 
We once again found that both \textit{C. opuntiae} and Other ants are parasites when they are the only partner present (Figure \ref{fig:BetweenLambdaMeans}a).
Diversity is beneficial through a sampling effect in this simulation (Figure \ref{fig:BetweenLambdaMeans}b).

\begin{figure}
\includegraphics[width=0.61\linewidth]{Figures/lambda_means_between.png}
\includegraphics[width=0.39\linewidth]{Figures/Lambda_Num_Partners_between.png}
\caption{Panel a) shows the mean values of the estimated $\lambda_{S}$ (filled in circles) for each simulated combination of ant partners. The letters above the points correspond to what ant partners are present (V = Vacant, C = \textit{C. opuntiae}, L = \textit{L. apiculatum}, O = other). Panel b) shows the mean values of the estimated $\lambda_{S}$ for }
\label{fig:BetweenLambdaMeans}
\end{figure}

\section*{Appendix D: Posterior Checks and Model Validation}
